---
title: "Initial Exploratory Data Analysis for STAT 551 Final Project"
author: "Kyle Maher"
date: today
format:
  html:
    self-contained: true
# documentclass: article
execute:
  warning: false
  message: false
---

# Setup
```{r}
library(corrplot)
library(dplyr)
library(ggplot2)
```

# Data
```{r}
data <- read.csv("FP_Data.csv")
df <- as.data.frame(data)
```

# EDA
```{r}
names(df)
```

## Correlations
```{r}
corrplot(cor(select(df, -x8)))
```

## Pairplot
```{r}
pairs(select(df, -x8))
```

## X1
```{r}
df %>%
  ggplot(aes(x = x1, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X2
```{r}
df %>%
  ggplot(aes(x = x2, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X3
```{r}
df %>%
  ggplot(aes(x = x3, y = y, color = x8)) +
  geom_point()
```

Slight positive correlation.

## X4
```{r}
df %>%
  ggplot(aes(x = x4, y = y, color = x8)) +
  geom_point()
```

Strong positive correlation.

## X5
```{r}
df %>%
  ggplot(aes(x = x5, y = y, color = x8)) +
  geom_point()
```

Strong negative correlation.

## X6
```{r}
df %>%
  ggplot(aes(x = x6, y = y, color = x8)) +
  geom_point()
```

Slight positive correlation.

## X7
```{r}
df %>%
  ggplot(aes(x = x7, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X8
```{r}
df %>%
  ggplot(aes(x = x8, y = y, fill = x8)) +
  geom_boxplot() +
  theme(
    legend.position = "none"
  )
```

As the classification increments (A-D) y tends to decrease.

- Outlier in A at 100
- Outlier in D at 0

## y
```{r}
df %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 15, aes(y = after_stat(density))) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(df$y), sd = sd(df$y)),
    color = "red",
    linewidth = 1
  )
``` 

```{r}
df %>%
  ggplot(aes(x = y, fill = x8)) +
  geom_histogram(bins = 10) +
  facet_wrap(~x8)
```

- A right skewed
- D left skewed


# Checking Multicolinearity

## x3 & x4
```{r}
df %>%
  ggplot(aes(x = x3, y = x4, color = x8)) +
  geom_point()
```

## x4 & x5
```{r}
df %>%
  ggplot(aes(x = x4, y = x5, color = x8)) +
  geom_point()
```

## x6 & x7
```{r}
df %>%
  ggplot(aes(x = x6, y = x7, color = x8)) +
  geom_point()
```

- Wacky $cos(x)$ function here