---
title: "Initial Exploratory Data Analysis for STAT 551 Final Project"
author: "Kyle Maher"
date: today
format:
  html:
    self-contained: true
# documentclass: article
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

# Setup
```{r}
library(corrplot)
library(dplyr)
library(ggplot2)
```

# Data
```{r}
data <- read.csv("FP_Data.csv")
df <- as.data.frame(data)
```

# EDA
```{r}
names(df)
```

## Correlations
```{r}
corrplot(cor(select(df, -x8)))
```

## Pairplot
```{r}
pairs(select(df, -x8))
```

## X1
```{r}
df %>%
  ggplot(aes(x = x1, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X2
```{r}
df %>%
  ggplot(aes(x = x2, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X3
```{r}
df %>%
  ggplot(aes(x = x3, y = y, color = x8)) +
  geom_point()
```

Slight positive correlation.

## X4
```{r}
df %>%
  ggplot(aes(x = x4, y = y, color = x8)) +
  geom_point()
```

Strong positive correlation.

## X5
```{r}
df %>%
  ggplot(aes(x = x5, y = y, color = x8)) +
  geom_point()
```

Strong negative correlation.

## X6
```{r}
df %>%
  ggplot(aes(x = x6, y = y, color = x8)) +
  geom_point()
```

Slight positive correlation.

## X7
```{r}
df %>%
  ggplot(aes(x = x7, y = y, color = x8)) +
  geom_point()
```

Pretty much random noise.

## X8
```{r}
df %>%
  ggplot(aes(x = x8, y = y, fill = x8)) +
  geom_boxplot() +
  theme(
    legend.position = "none"
  )
```

As the classification increments (A-D) y tends to decrease.

- Outlier in A at 100
- Outlier in D at 0

## y
```{r}
df %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 15, aes(y = after_stat(density))) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(df$y), sd = sd(df$y)),
    color = "red",
    linewidth = 1
  )
``` 

```{r}
df %>%
  ggplot(aes(x = y, fill = x8)) +
  geom_histogram(bins = 10) +
  facet_wrap(~x8)
```

- A right skewed
- D left skewed

```{r}
xtabs(~x8, df)
```

### Grouped Histograms

```{r}
df %>%
  filter(x8 == "A" | x8 == "C") %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 10, fill = "steelblue") +
  labs(
    title = ("Distribution of y when x8 is B or D")
  )
```

```{r}
df %>%
  filter(x8 == "B" | x8 == "D") %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 10, fill = "steelblue") +
  labs(
    title = "Distribution of y when x8 is B or D"
  )
```


# Checking Multicolinearity

## x3 & x4
```{r}
df %>%
  ggplot(aes(x = x3, y = x4, color = x8)) +
  geom_point()
```

## x4 & x5
```{r}
df %>%
  ggplot(aes(x = x4, y = x5, color = x8)) +
  geom_point()
```

## x6 & x7
```{r}
df %>%
  ggplot(aes(x = x6, y = x7, color = x8)) +
  geom_point()
```

- Wacky $cos(x)$ function here

# Baseline Model
```{r}
baseline_model <- lm(y ~ ., data = df)
summary(baseline_model)
```

- R-squared of 0.6672, Adj. R-squared of 0.6496

```{r}
smaller_lm <- lm(y ~ x4 + x8 + x7, data = df)
summary(smaller_lm)
```

- We can get R-squared of 0.6489 with just three predictors

## PCA

```{r}
df_onehot <- df |> mutate(xA = ifelse(x8 == "A", 1, 0),
                          xB = ifelse(x8 == "B", 1, 0),
                          xC = ifelse(x8 == "C", 1, 0),
                          xD = ifelse(x8 == "D", 1, 0)) |> select(-x8)
df_onehot_no_response <- df_onehot |> select(-y)

pca_results <- prcomp(df_onehot_no_response, scale = TRUE)
# biplot(pca_results, scale = 0)

summary(lm(df$y ~ pca_results$x[, c(1, 2, 7, 9, 11)]))
```

- R-squared of 0.6538 with five predictors

```{r}
library(pls)

pcr_model <- pcr(y ~ ., data = df_onehot, validation = "CV")
summary(pcr_model)

summary(lm(y ~ ., data = df_onehot))
```

- PCR with all components slightly beats standard linear regression (0.58 accuracy difference)