---
title: "Final Report for STAT 451/551"
author: "Jack Barney, Kyle Maher, Max Donelan, & Tate Wietfeld"
date: 2026-05-05
date-format: "[Spring] YYYY"
format:
  pdf:
    self-contained: true
    documentclass: scrreprt
    mainfont: "Times New Roman"
    fontsize: 12pt
    linestretch: 1.5
    geometry:
      - margin=1in
    pdf-engine: xelatex
execute:
  warning: false
  message: false
  echo: false
editor_options: 
  chunk_output_type: console
header-includes:
  - \graphicspath{{../../}}
---

<!-- 
Frontpage. Title, Contributors; Semester & Year. 

Introduction: A brief intro, restate the general objective, nature of variables in the dataset. At the end of this section, include a sentence how the rest of your report is organized.  

Descriptive analyses: Includes graphical and numerical descriptive statistics, preliminary discussions regarding patterns in the data.
Fitting and evaluating various predictive models. Continuous (y) and dichotomized (ybin) cases.   

Concluding remarks. 100 - 120 words. 

Sources. Textbook, research papers, internet, etc. Use APA style.

Appendix. A brief explanation of the programs used. Supplemental or extended results as needed.  
-->

# Introduction

The given FP data contains $200$ observations of eight predictors labeld $x1$ through $x8$, and the resulting response parameter $y$. Additional consideration was given to predicting $y\_bin$ which is $1$ if $y$ is greater than or equal to $70$ and $0$ otherwise. $x1$ through $x7$ are continuous while $x8$ is categorical with four groups $A, B, C,$ and $D$. No context surrounding the predictors or response parameters was given. This report is segmented into initial Exploratory Data Analysis, followed by the Regression Models, then the Classification Models, and concludes with some overall observations. All written code can be found in the appendix.

<!-- SETUP: All Libraries -->
```{r}
library(dplyr)         # Data Manipulation
library(ggplot2)       # Plotting
library(pls)           # PCR
library(corrplot)      # Correlations Plot (function in pls masks)
library(randomForest)  # Random Forest
library(gbm)           # Boosted Trees
library(class)         # knn
library(glmnet)        # Regularized Logistic
library(mgcv)          # Logistic GAM

set.seed(2800)
```

# Exploratory Data Analysis

```{r}
data <- read.csv("data/FP_Data.csv")
df <- as.data.frame(data)
df <- df %>%
  mutate(x8 = as.factor(x8))
```

## Continuous $y$

### Correlations
```{r}
#| fig-height: 4
#| fig-width: 4
#| fig-cap: "Correlations of Continuous Parameters"

corrplot(cor(select(df, -x8)))
```

### Pairplot
```{r}
#| fig-height: 6
#| fig-width: 6
#| fig-cap: "Scatterplots of Continuous Parameters"

pairs(select(df, -x8))
```

### y
```{r}
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Histogram of y"

df %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 15, aes(y = after_stat(density))) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(df$y), sd = sd(df$y)),
    color = "red",
    linewidth = 1
  )
``` 

```{r}
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Histograms of y by x8"

df %>%
  ggplot(aes(x = y, fill = x8)) +
  geom_histogram(bins = 10) +
  facet_wrap(~x8)
```

- A right skewed
- D left skewed

## Dichotomized $y$


# Regression Models

<!-- ## Setup -->
```{r}
set.seed(2800)
k <- 10
# Stratifiy by x8 group
df <- df %>%
  group_by(x8) %>%
  mutate(fold = sample(rep(1:k, length.out = n()))) %>%
  ungroup() %>%
  mutate(x8 = as.factor(x8))  # For gbm() (BostedTrees)

# Output Cross Validation Split for Python Models
write.csv(df, "src/04-cv/10-fold-cv.csv", row.names = FALSE)

model_names <- c("BaselineLM", "SmallLM", "PCR", "RandomForest", "BoostedTrees")
model_count <- length(model_names)
cv_stats <- data.frame(
  fold = rep(1:k, model_count),
  model = rep(model_names, each = k),
  rmse = NA,
  mae = NA
)
# Root Mean Squared Error
rmse <- function(actual, predicted) {
  return(sqrt(mean((actual - predicted)^2)))
}
# Mean Absolute Error
mae <- function(actual, predicted) {
  return(mean(abs(actual - predicted)))
}
```

<!-- ## Running the Cross Validation -->
```{r}
for (i in 1:k) {
  train_df <- filter(df, fold != i) %>%
    select(-fold)
  test_df <- filter(df, fold == i) %>%
    select(-fold)
  
  # Baseline LM
  baseline_lm <- lm(y ~ ., data = train_df)
  pred <- predict(baseline_lm, newdata = test_df)
  cv_stats[cv_stats$model == "BaselineLM", "rmse"][i] <- rmse(test_df$y, pred)
  cv_stats[cv_stats$model == "BaselineLM", "mae"][i] <- mae(test_df$y, pred)

  # Small LM
  small_lm <- lm(y ~ x4 + x8 + x7, data = train_df)
  pred <- predict(small_lm, newdata = test_df)
  cv_stats[cv_stats$model == "SmallLM", "rmse"][i] <- rmse(test_df$y, pred)
  cv_stats[cv_stats$model == "SmallLM", "mae"][i] <- mae(test_df$y, pred)

  # PCR
  train_df_onehot <- train_df %>% 
    mutate(
      xA = ifelse(x8 == "A", 1, 0),
      xB = ifelse(x8 == "B", 1, 0),
      xC = ifelse(x8 == "C", 1, 0),
      xD = ifelse(x8 == "D", 1, 0)
      ) %>% 
    select(-x8)
  test_df_onehot <- test_df  %>% 
    mutate(
      xA = ifelse(x8 == "A", 1, 0),
      xB = ifelse(x8 == "B", 1, 0),
      xC = ifelse(x8 == "C", 1, 0),
      xD = ifelse(x8 == "D", 1, 0)
      ) %>% 
    select(-x8)
  pcr_model <- pcr(y ~ ., data = train_df_onehot, validation = "CV")
  num_components <- selectNcomp(pcr_model, method = "onesigma")
  pred <- predict(pcr_model, newdata = test_df_onehot)[, , num_components]
  cv_stats[cv_stats$model == "PCR", "rmse"][i] <- rmse(test_df_onehot$y, pred)
  cv_stats[cv_stats$model == "PCR", "mae"][i] <- mae(test_df_onehot$y, pred)

  # Random Forest
  rf_model <- randomForest(y ~ ., data = train_df, mtry = 6, importance = TRUE)
  pred <- predict(rf_model, newdata = test_df)
  cv_stats[cv_stats$model == "RandomForest", "rmse"][i] <- rmse(test_df$y, pred)
  cv_stats[cv_stats$model == "RandomForest", "mae"][i] <- mae(test_df$y, pred)

  # Boosted Trees
  bt_model <- gbm(
    y ~ .,
    data = train_df,
    distribution = "gaussian", 
    n.trees = 5000,
    interaction.depth = 4
  )
  pred <- predict(bt_model, newdata = test_df, n.trees = 5000)
  cv_stats[cv_stats$model == "BoostedTrees", "rmse"][i] <- rmse(test_df$y, pred)
  cv_stats[cv_stats$model == "BoostedTrees", "mae"][i] <- mae(test_df$y, pred)
}
```

<!-- ## Add Python Neural Network Results -->
```{r}
nn_cv <- read.csv("src/04-cv/reg_nn_cv_results.csv")
cv_stats <- bind_rows(cv_stats, nn_cv)
```

## Plotting the 10-Fold Cross Validation Metrics

```{r}
#| fig-cap: "RMSE by Model for Each Fold"

cv_stats %>%
  ggplot(aes(x = model, y = rmse, fill = model)) +
  geom_boxplot(alpha = .8, outlier.shape = NA) +
  geom_jitter(width = .2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "RMSE",
    title = "Root Mean Squared Error by Model"
  )
```

```{r}
#| fig-cap: "RMSE by Model and by Fold"

cv_stats %>%
  ggplot(aes(x = model, y = rmse, color = as.factor(fold), group = fold)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "RMSE",
    title = "RMSE by Model and by Fold"
  )
```

```{r}
#| fig-cap: "MAE by Model for Each Fold"

cv_stats %>%
  ggplot(aes(x = model, y = mae, fill = model)) +
  geom_boxplot(alpha = .8, outlier.shape = NA) +
  geom_jitter(width = .2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "MAE",
    title = "Mean Absolute Error by Model"
  )
```


# Classification Models

```{r}
data <- read.csv("data/FP_Data.csv")
df <- as.data.frame(data)
df <- df %>%
  mutate(
    y_bin = ifelse(y >= 70, 1, 0),
    x8 = as.factor(x8)
  ) %>%
  select(-y)
```

<!-- ## Setup -->
```{r}
set.seed(2800)
k <- 10
# Stratifiy by x8 group
df <- df %>%
  group_by(x8) %>%
  mutate(fold = sample(rep(1:k, length.out = n()))) %>%
  ungroup()

model_names <- c(
  "BaselineLogistic", 
  "SmallLogistic", 
  "RegularizedLogisitc", 
  "LogisticGAM", 
  "KNN"
)

model_count <- length(model_names)
cv_stats <- data.frame(
  fold = rep(1:k, model_count),
  model = rep(model_names, each = k),
  accuracy = NA,
  fpr = NA,
  fnr = NA
)
# Accuracy
accuracy <- function(actual, predicted) {
  return(mean(actual == predicted))
}

# False Positive Rate
fpr <- function(actual, predicted) {
  return(sum(predicted == 1 & actual == 0) / sum(actual == 0))
}

# False Negative Rate
fnr <- function(actual, predicted) {
  return(sum(predicted == 0 & actual == 1) / sum(actual == 1))
}
```

<!-- ## Running the Cross Validation -->
```{r}
for (i in 1:k) {
  train_df <- filter(df, fold != i) %>%
    select(-fold)
  test_df <- filter(df, fold == i) %>%
    select(-fold)
  
  # Baseline Logistic
  baseline_logistic <- glm(y_bin ~ ., data = train_df, family = "binomial")
  pred_prob <- predict(baseline_logistic, newdata = test_df, type = "response")
  pred <- ifelse(pred_prob >= .5, 1, 0)
  cv_stats[cv_stats$model == "BaselineLogistic", "accuracy"][i] <- accuracy(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "BaselineLogistic", "fpr"][i] <- fpr(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "BaselineLogistic", "fnr"][i] <- fnr(pred, test_df$y_bin)

  # Small Logistic
  small_logistic <- glm(y_bin ~ x1 + x3 + x5 + x7 + x8, data = train_df, family = "binomial")
  pred_prob <- predict(small_logistic, newdata = test_df, type = "response")
  pred <- ifelse(pred_prob >= .5, 1, 0)
  cv_stats[cv_stats$model == "SmallLogistic", "accuracy"][i] <- accuracy(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "SmallLogistic", "fpr"][i] <- fpr(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "SmallLogistic", "fnr"][i] <- fnr(pred, test_df$y_bin)

  # Regularized Logistic
  train_mat <- as.matrix(select(train_df, -y_bin))
  test_mat <- as.matrix(select(test_df, -y_bin))
  cv <- cv.glmnet(train_mat, train_df$y_bin, family = "binomial", alpha = 1)
  lambda <- cv$lambda.min
  reg_logistic <- glmnet(train_mat, train_df$y_bin, family = "binomial", alpha = 1, lambda = lambda)
  predict_probs <- predict(reg_logistic, newx = test_mat, s = lambda, type = "response")
  pred <- ifelse(predict_probs >=.5, 1, 0)
  cv_stats[cv_stats$model == "RegularizedLogisitc", "accuracy"][i] <- accuracy(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "RegularizedLogisitc", "fpr"][i] <- fpr(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "RegularizedLogisitc", "fnr"][i] <- fnr(pred, test_df$y_bin)

  # Logistic GAM
  logistic_gam <- gam(y_bin ~ s(x5) + s(x7), family = "binomial", data = train_df, method = "REML")
  predict_probs <- predict(logistic_gam, newdata = test_df, type = "response")
  pred <- ifelse(predict_probs >=.5, 1, 0)
  cv_stats[cv_stats$model == "LogisticGAM", "accuracy"][i] <- accuracy(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "LogisticGAM", "fpr"][i] <- fpr(pred, test_df$y_bin)
  cv_stats[cv_stats$model == "LogisticGAM", "fnr"][i] <- fnr(pred, test_df$y_bin)

  # KNN
  train_df_onehot <- train_df %>% 
    mutate(
      x1 = scale(x1),
      x2 = scale(x2),
      x3 = scale(x3),
      x4 = scale(x4),
      x5 = scale(x5),
      x6 = scale(x6),
      x7 = scale(x7),
      xA = ifelse(x8 == "A", 1, 0),
      xB = ifelse(x8 == "B", 1, 0),
      xC = ifelse(x8 == "C", 1, 0),
      xD = ifelse(x8 == "D", 1, 0)
      ) %>% 
    select(-x8)
  test_df_onehot <- test_df  %>% 
    mutate(
      x1 = scale(x1),
      x2 = scale(x2),
      x3 = scale(x3),
      x4 = scale(x4),
      x5 = scale(x5),
      x6 = scale(x6),
      x7 = scale(x7),
      xA = ifelse(x8 == "A", 1, 0),
      xB = ifelse(x8 == "B", 1, 0),
      xC = ifelse(x8 == "C", 1, 0),
      xD = ifelse(x8 == "D", 1, 0)
      ) %>% 
    select(-x8)

  k_vals <- seq(1, 21, by = 2)
  k_accuracy <- data.frame(k = k_vals, accuracy = NA)

  for (k in k_vals) {
    pred <- knn(
      select(train_df_onehot, -y_bin), 
      select(test_df_onehot, -y_bin), 
      as.factor(train_df_onehot$y_bin), 
      k = k
    )
    k_accuracy[k_accuracy$k == k, "accuracy"] <- mean(pred == test_df_onehot$y_bin)
  }

  best_k <- k_accuracy$k[which.max(k_accuracy$accuracy)]
  pred <- knn(
    select(train_df_onehot, -y_bin), 
    select(test_df_onehot, -y_bin), 
    as.factor(train_df_onehot$y_bin), 
    k = best_k
  )

  cv_stats[cv_stats$model == "KNN", "accuracy"][i] <- accuracy(pred, test_df_onehot$y_bin)
  cv_stats[cv_stats$model == "KNN", "fpr"][i] <- fpr(pred, test_df_onehot$y_bin)
  cv_stats[cv_stats$model == "KNN", "fnr"][i] <- fnr(pred, test_df_onehot$y_bin)

}
```

<!-- ## Adding the Python Neural Network Results -->

```{r}
nn_cv <- read.csv("src/04-cv/class_nn_cv_results.csv")
cv_stats <- bind_rows(cv_stats, nn_cv)
```

## Plotting the 10-Fold Cross Validation Metrics

```{r}
#| fig-cap: "Accuracy by Model for Each Fold"

cv_stats %>%
  ggplot(aes(x = model, y = accuracy, fill = model)) +
  geom_boxplot(alpha = .8, outlier.shape = NA) +
  geom_jitter(width = .2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "Accuracy",
    title = "Accuracy by Model"
  )
```

```{r}
#| fig-cap: "False Positive Rate by Model for Each Fold"

cv_stats %>%
  ggplot(aes(x = model, y = fpr, fill = model)) +
  geom_boxplot(alpha = .8, outlier.shape = NA) +
  geom_jitter(width = .2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "False Positive Rate",
    title = "False Positive Rate by Model"
  )
```

```{r}
#| fig-cap: "False Negative Rate by Model for Each Fold"

cv_stats %>%
  ggplot(aes(x = model, y = fnr, fill = model)) +
  geom_boxplot(alpha = .8, outlier.shape = NA) +
  geom_jitter(width = .2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "False Negative Rate",
    title = "False Negative Rate by Model"
  )
```

```{r}
#| fig-cap: "Accuracy by Model and by Fold"

cv_stats %>%
  ggplot(aes(x = model, y = accuracy, color = as.factor(fold), group = fold)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  ) +
  labs(
    x = "Model",
    y = "Accuracy",
    title = "Accuracy by Model and by Fold"
  )
```